Towards a Reinforcement Learning approach for Aggregate Computing

### 2021-11-29 paper-2021-woa-ac-ml.tex
minors

\section{Introduction}
Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
 promote a vision of artificially designed Collective Adaptive Systems (CASs)~\cite{DBLP:conf/huc/Ferscha15},
 namely collections of agents coordinating and adapting %as wholes 
 to pursue joint goals in changing environments.
% due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
%
%CASs are characterised by 
%\begin{itemize}
%\item a global order that emerges from local node interactions, 
%\item  decentralised control, and 
%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
%\end{itemize}
%
CAS are notoriously hard to design and control 
 %due to as part of complex systems
 as globally desired behaviour
 has to emerge from local behaviours and interactions. 
%
For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.). 
%
%In this case, a sort of hive mind arises 
% %-- also referred to as swarm intelligence --  
% born from local interactions between agents. 
%
Specifically, such approaches tend to exploit 
 \emph{self-organisation}, 
 i.e., processes by which global order emerges 
 from continuous interactions among simple entities.

Initially, researchers had tried to build systems that show these properties in computing systems by mimickng natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to achieve the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different elements not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt to the non stationarity due to node movements.
%
Ideally, we would like to continue developing our applications using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden of defining the right collective specifications to developers.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

Instead, reward function might be dependent on the learning technique, time, and problem. 
 Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
$$
\textit{R}_{global} := \psi({\theta})
$$
If we use an online mode, it needs to be defined as a function to the local node context:
$$
\textit{R}_{local} := \psi(\textit{Context})
$$
Finally, actions influence the current aggregate program output, changing the program evaluation as:
$$
f_{rl} : \textit{Context} \rightarrow \lambda(\textit{Output}, \textit{Action})
$$
As remarked before, the policy $\pi$ needs to be distributed, so that each agent can choose the action that will be performed:
$$
\textit{Action} \leftarrow \pi(K)
$$

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
 and effectiveness of the AC framework.
In particular, we explore the combination of AC with Hysteretic Q-Learning.
The application of RL to this kind of setting is a largely unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
 research area. There are a variety of challenges, such as non-stationarity,  partial environment observability, and non-bounded system size.
In a case study, we have empirically showed the effectiveness of RL by showing a speed up of the convergence time of
 aggregate program we took into consideration, namely the hop count.
Furthermore, we have shown that the learnt program can easily scale up to system size.

>>>>

\section{Introduction}
Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
 promote a vision of artificially designed Collective Adaptive Systems (CASs)~\cite{DBLP:conf/huc/Ferscha15},
 namely collections of agents coordinating and adapting %as wholes 
 to pursue joint goals in changing environments.
% due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
%
%CASs are characterised by 
%\begin{itemize}
%\item a global order that emerges from local node interactions, 
%\item  decentralised control, and 
%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
%\end{itemize}
%
CAS are notoriously hard to design and control 
 %due to as part of complex systems
 as globally desired behaviour
 has to emerge from local behaviours and interactions. 
%
For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, beehives, flocks of birds etc.). 
%
%In this case, a sort of hive mind arises 
% %-- also referred to as swarm intelligence --  
% born from local interactions between agents. 
%
Specifically, such approaches tend to exploit 
 \emph{self-organisation}, 
 i.e., processes by which global order emerges 
 from continuous interactions among simple entities.

Initially, researchers had tried to build systems that show these properties in computing systems by mimicking natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to achieve the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different elements not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt to the non-stationarity due to node movements.
%
Ideally, we would like to continue developing our applications using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden of defining the right collective specifications to developers.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

Instead, the reward function might be dependent on the learning technique, time, and problem. 
 Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
$$
\textit{R}_{global} := \psi({\theta})
$$
If we use an online mode, it needs to be defined as a function to the local node context:
$$
\textit{R}_{local} := \psi(\textit{Context})
$$
Finally, actions influence the current aggregate program output, changing the program evaluation as:
$$
f_{rl} : \textit{Context} \rightarrow \lambda(\textit{Output}, \textit{Action})
$$
As remarked before, the policy $\pi$ needs to be distributed, so that each agent can choose the action that will be performed:
$$
\textit{Action} \leftarrow \pi(K)
$$

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
 and effectiveness of the AC framework.
In particular, we explore the combination of AC with Hysteretic Q-Learning.
The application of RL to this kind of setting is a largely unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
 research area. There are a variety of challenges, such as non-stationarity,  partial environment observability, and non-bounded system size.
In a case study, we have empirically shown the effectiveness of RL by showing a speedup of the convergence time of
 the aggregate program we took into consideration, namely the hop count.
Furthermore, we have shown that the learnt program can easily scale up to system size.
